---
title: Representing 'omics relationships with _IndexedRelations_
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
date: "Revised: 10 April 2019"
output:
  BiocStyle::html_document:
    toc_float: true
package: IndexedRelations
vignette: >
  %\VignetteIndexEntry{Representing 'omics relationships} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Overview

The `r Biocpkg("IndexedRelations")` package implements the `IndexedRelations` class for representing 'omics relationships.
This can be used for physical interactions, regulatory relationships or other associations involving any number of features.
The `IndexedRelations` class is primarily intended as a base class from which concrete subclasses can be derived for specific contexts.
For example, the `r Biocpkg("GenomicInteractions")` package derives an subclass to represent physical interactions between genomic intervals.

# The `IndexedRelations` class

Each `IndexedRelations` object holds relationships between "partner" features.
Partners can be any instance of a `Vector`-like class such as `GenomicRanges` and `IRanges`^[See `r Biocpkg("S4Vectors")` for details.].
In a single `IndexedRelations` object, each entry is a relationship that involves the same number and type of partners.

To illustrate, let's make up a `GenomicRanges` and an `IRanges` object.
The former might represent a chromosomal interval while the latter might represent, say, an interval on an exogenous sequence like a plasmid.

```{r}
# Making up features.
library(GenomicRanges)
gr <- GRanges(sample(1:21, 100, replace=TRUE),
    IRanges(sample(1e8, 100, replace=TRUE), width=10))
gr

ir <- IRanges(sample(1000, 20, replace=TRUE), width=5)
ir
```

Now, assume that there are multiple relationships between individual entries of these two objects.
(Let's pretend that the parts of the plasmid come into contact with the chromosomes at various points.)
The `partner.G` vector refers to elements of `gr` while `partner.I` refers to elements of `ir`.
The first element of `partner.G` is in a relationship with the first element of `partner.I`, and so on.

```{r}
# Making up relationships.
partner.G <- sample(length(gr), 10000, replace=TRUE)
partner.I <- sample(length(ir), 10000, replace=TRUE)

# Parallel entries across vectors are in a relationship:
gr[partner.G[1]]
ir[partner.I[1]]
```

It is then straightforward to construct an `IndexedRelations` object.
While this example only uses two partnering features, the class can support any number of partners in a single object.

```{r}
# Easy but inefficient:
library(IndexedRelations)
rel <- IndexedRelations(list(gr[partner.G], ir[partner.I]))

# More work but more efficient:
rel <- IndexedRelations(list(partner.G, partner.I), list(gr, ir))
rel
```

The key feature of the `IndexedRelations` class is that it does not actually store the partnering features explicitly.
Rather, the relationships are internally represented by indices that point to sets of features.
This avoids redundant representation of the same features when large numbers of relationships are to be stored,
reducing memory use and improving the efficiency of algorithms that operate on these relationships.

# Getters and setters

Use the `partner()` method to extract the partners:

```{r}
# Extract partnering features:
partner(rel, 1)
partner(rel, 2)

# Extract partnering indices only:
head(partner(rel, 1, id=TRUE))
head(partner(rel, 2, id=TRUE))
```

Use the `featureSets()` method to extract the feature sets:

```{r}
featureSets(rel)[[1]]
featureSets(rel)[[2]]
```

The `IndexedRelations` object behaves like a vector.
It has `length`, `names` and can be subsetted and combined.

```{r}
length(rel)
length(rel[1:10,])
length(c(rel, rel))
```

You can also set the partners or features on an existing `IndexedRelations` object:

```{r}
# Set partner with features.
partner(rel, 2) <- ir[sample(length(ir), length(rel), replace=TRUE)]

# Set partner with indices.
partner(rel, 2, id=TRUE) <- sample(length(ir), length(rel), replace=TRUE)

# Modify the first feature set.
featureSets(rel)[[1]] <- resize(featureSets(rel)[[1]], 20)
```

In most cases, an `IndexedRelations` object will behave "as if" it were an object containing the partnering features explicitly.
Users do not have to worry about the specifics of the index-based representation unless they are specifically manipulating it.

# Conversions

The `IndexedRelations` object can be converted to and from some similar classes.
The most obvious of these is the `Pairs` class from `r Biocpkg("S4Vectors")`^[Of course, this only works if the `IndexedRelations` object has two partners!]:

```{r}
makePairsFromIndexedRelations(rel)

p <- Pairs(gr[1:10], ir[1:10])
as(p, "IndexedRelations")
```

Another option is to convert to and from a `DataFrame` object.
This provides a quick and general method to "realize" the indices into the partnering features.

```{r}
as(rel, "DataFrame")
```

# Session info

```{r}
sessionInfo()
```
